---
title: "Introduction to MBbioeq"
subtitle: "Model-Based Bioequivalence Analysis for Parallel-Arm Studies"
author: "Eden BETE, Célia BENGUENNA"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to MBbioeq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

**MBbioeq** implements *model-based bioequivalence (BE) analysis* for
parallel-arm study designs using nonlinear mixed-effects pharmacokinetic (PK)
modelling.  Classical NCA-based (non-compartmental analysis) BE methods
require a sufficiently rich sampling design per subject; model-based methods
can recover reliable treatment comparisons even from sparse designs because
information is pooled across subjects through the population model.

The package estimates a one-compartment PK model with the SAEM algorithm
(**saemix** package) and then tests whether the Test (T) and Reference (R)
formulations are bioequivalent on the AUC and C~max~ endpoints using two
complementary procedures:

| Procedure | Key idea |
|-----------|----------|
| **TOST** — Two One-Sided Tests | 90 % CI for the T/R ratio must lie within [0.80, 1.25] |
| **BOT** — Bayesian-Optimal Test | Observed \|β̂\| must be ≤ the α-quantile of a folded-normal under H₀ |

Both tests accept the same model fit, so the computational cost is the same.

# Installation

```{r install, eval = FALSE}
# From a local source directory:
devtools::install("path/to/MBbioeq")

# Or once submitted to CRAN:
install.packages("MBbioeq")
```

```{r load}
library(MBbioeq)
```

# Dataset format

## Required columns

The dataset must be a plain text file (space-, tab-, semicolon-, or
comma-separated) with a header row.  The following columns are required
(exact names are case-insensitive; accepted synonyms are listed):

| Canonical name | Accepted synonyms | Description |
|----------------|-------------------|-------------|
| `Id` | `subject`, `subj`, `sid` | Subject identifier |
| `Time` | `t`, `hours`, `hr` | Sampling time (hours) |
| `Concentration` | `conc`, `dv`, `y`, `obs` | Observed concentration (mg/L) |
| `Treatment` | `treat`, `trt`, `tr`, `group`, `arm` | `"T"` = Test, `"R"` = Reference |
| `Dose` | `amt` | Administered dose (optional if carried forward) |

## Missing values

Missing concentrations (pre-dose samples, BLQ values, etc.) should be
coded as `.`, `NA`, or left empty.  Missing dose values are
**forward-filled** within each subject; at least one dose value per subject
must be present.

## Example

```
Id Dose Time Treatment Concentration
1  4    0    R         .
1  .    0.5  R         2.22
1  .    1.0  R         3.84
1  .    2.0  R         4.53
...
21 4    0    T         .
21 .    0.5  T         2.82
```

The bundled dataset `dataset_1.txt` is used throughout this vignette:

```{r data-path}
data_path <- system.file("extdata", "dataset_1.txt", package = "MBbioeq")
```

## Preparing the data

`prep_parallel_TR()` standardises column names, validates treatment coding,
and forward-fills missing doses.  It is called automatically inside
`run_MBTost()` and `run_MBbot()`, but you can also call it directly:

```{r prep, eval = FALSE}
raw <- read.table(data_path, header = TRUE)
tab <- prep_parallel_TR(raw)
head(tab)
#>   Id Dose Time Concentration Treat
#> 2  1    4  0.5         2.217     0
#> 3  1    4  1.0         3.840     0
#> ...
```

# Spaghetti plot

Before fitting any model it is good practice to inspect the raw profiles.
`plot_spaghetti()` overlays all individual concentration-time curves, coloured
by treatment group, with an optional group mean profile.

```{r spaghetti, eval = FALSE}
plot_spaghetti(data_path)
```

**How to read the plot:**

- **Coloured lines (transparent)** — individual subjects.  Variability within
  each group is immediately visible.
- **Bold line** — arithmetic mean profile of the group.  A higher or earlier
  peak for one treatment compared to the other is a first visual indication of
  a potential difference.
- If profiles overlap substantially, both T and R drive a similar exposure —
  a good sign for bioequivalence.

# TOST analysis

## Running the test

```{r tost-run, eval = FALSE}
psi0      <- c(ka = 1.5, V = 0.5, CL = 0.04)
omega_init <- c(0.2, 0.3, 0.2)   # diagonal random effects

res_tost <- run_MBTost(
  data_path   = data_path,
  model       = "1cpt",
  psi0        = psi0,
  omega_init  = omega_init,
  error_model = "combined",
  error_init  = c(0.1, 0.2)
)
```

A typical result looks like:

```
════════════════════════════════════════════════════════════
              TOST BIOEQUIVALENCE RESULTS
════════════════════════════════════════════════════════════
  Model: 1cpt  | n = 40  | k = 3  | δ = 0.2231

AUC RATIO (T/R):
  Estimate        : 1.0142
  90% CI (Normal) : [0.9287, 1.1073]  →  Bioequivalent
  90% CI (Gallant): [0.9198, 1.1195]  →  Bioequivalent

Cmax RATIO (T/R):
  Estimate        : 1.0289
  90% CI (Normal) : [0.9318, 1.1363]  →  Bioequivalent
  90% CI (Gallant): [0.9214, 1.1473]  →  Bioequivalent
════════════════════════════════════════════════════════════
```

## Forest plot

```{r tost-plot, eval = FALSE}
plot_tost(res_tost)
```

**How to read the forest plot:**

- The **green shaded rectangle** is the bioequivalence acceptance zone
  [0.80, 1.25].
- Each row shows one endpoint × inference method combination: the **diamond**
  is the geometric mean ratio (GMR), the **horizontal bar** is the 90 % CI,
  and tick marks show the CI endpoints.
- **Green** = CI fully inside the zone → Bioequivalent; **Red** = CI crosses a
  boundary → Not bioequivalent.
- The Gallant row always has a wider CI than the Normal row; it is the more
  conservative (and more reliable for small n) inference.

```{r tost-plot-show, eval = FALSE}
# Display only the Gallant-corrected CIs:
plot_tost(res_tost, method = "gallant")
```

# BOT analysis

## Running the test

```{r bot-run, eval = FALSE}
res_bot <- run_MBbot(
  data_path   = data_path,
  model       = "1cpt",
  psi0        = psi0,
  omega_init  = omega_init,
  error_model = "combined",
  error_init  = c(0.1, 0.2)
)
```

```
════════════════════════════════════════════════════════════
               BOT BIOEQUIVALENCE RESULTS
════════════════════════════════════════════════════════════
  Model: 1cpt  | n = 40  | k = 3  | δ = 0.2231

AUC RATIO (T/R):
  Estimate       : 1.0142
  |beta_hat|     : 0.0141
  Critical value : 0.1803
  Decision       : Bioequivalent

Cmax RATIO (T/R):
  Estimate       : 1.0289
  |beta_hat|     : 0.0285
  Critical value : 0.1926
  Decision       : Bioequivalent
════════════════════════════════════════════════════════════
```

## Folded-normal decision plot

```{r bot-plot, eval = FALSE}
plot_bot(res_bot)
```

**How to read the folded-normal plot:**

- The **bell-shaped curve** is the density of the folded-normal distribution
  $FN(\delta, \hat\sigma^2)$ — this is the worst-case null distribution of
  $|\hat\beta|$ when the true effect is exactly at the equivalence boundary
  $\delta$.
- The **dashed red line** ($c^*$) is the $\alpha$-quantile of that
  distribution: by construction, $P(|\hat\beta| \le c^* \mid |\beta| = \delta)
  = \alpha$.
- The **solid blue line** is the observed $|\hat\beta|$.
- **Green region** ($[0, c^*]$) — the acceptance region; if the blue line falls
  here, bioequivalence is concluded.
- **Red region** ($(c^*, \infty)$) — the rejection region.
- The smaller $|\hat\beta|$ relative to $c^*$, the stronger the evidence for
  bioequivalence.

# PK model choice

Two models are available:

| Model | Parameters | Use when |
|-------|-----------|---------|
| `"1cpt"` | ka, V, CL | Simple oral absorption, no lag |
| `"1cpt_tlag"` | ka, V, CL, tlag | Delayed absorption (e.g. enteric coating) |

```{r tlag-example, eval = FALSE}
res_tlag <- run_MBTost(
  data_path   = data_path,
  model       = "1cpt_tlag",
  psi0        = c(ka = 1.5, V = 0.5, CL = 0.04, tlag = 0.5),
  omega_init  = c(0.2, 0.3, 0.2, 0.1),
  error_model = "combined"
)
plot_tost(res_tlag)
```

# Advanced options

## Covariance structure

Random effects can be assumed independent (`"diagonal"`, default) or fully
correlated (`"full"`).  A custom binary matrix can also be passed:

```{r cov-example, eval = FALSE}
# Full covariance matrix (all off-diagonals estimated)
run_MBTost(data_path, model = "1cpt", covariance_structure = "full")

# Custom: only CL-V covariance
omega_block <- matrix(c(1, 0, 0,
                        0, 1, 1,
                        0, 1, 1), nrow = 3)
run_MBTost(data_path, model = "1cpt", covariance_structure = omega_block)
```

## Equivalence margin

The default margin $\delta = \log(1.25)$ corresponds to the regulatory
[0.80, 1.25] acceptance range.  Narrow therapeutic index drugs may use a
tighter margin:

```{r delta-example, eval = FALSE}
run_MBTost(data_path, model = "1cpt", delta = log(1.111))  # 90–111 % range
```

## Significance level

`alpha` sets the one-sided significance level; the two-sided confidence level
is $1 - 2\alpha$.  The default `alpha = 0.05` yields a 90 % CI.

```{r alpha-example, eval = FALSE}
run_MBTost(data_path, model = "1cpt", alpha = 0.10)  # 80 % CI (more liberal)
```

# Saving plots

All plot functions return their data invisibly, so you can capture them with
standard R graphics devices:

```{r save-plot, eval = FALSE}
pdf("tost_result.pdf", width = 8, height = 5)
plot_tost(res_tost)
dev.off()

png("spaghetti.png", width = 800, height = 500)
plot_spaghetti(data_path)
dev.off()
```

# Session information

```{r session}
sessionInfo()
```
